---
title: "02 - Data Exxploration"
author: "Humbert Costas, Sergi Puigventos Ventura & Daniel Da Graça"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, results='hide'}
knitr::opts_chunk$set(echo = TRUE)
# Packages required for the project
required_packages <- c('rstudioapi', 'fs', 'stringr', 'knitr', 'randomForest', 'readr', 'caret', 'e1071', 'tidyverse', 'ggplot2', 'dplyr', 'corrplot')
# Installation of all required packages not already installed
install.packages(setdiff(required_packages, rownames(installed.packages())))
# Loading required packages
library('rstudioapi', warn.conflicts = FALSE)
library('fs', warn.conflicts = FALSE)
library('stringr', warn.conflicts = FALSE)
library('randomForest', warn.conflicts = FALSE)
library('readr', warn.conflicts = FALSE)
library('caret', warn.conflicts = FALSE)
library('e1071', warn.conflicts = FALSE)
library('tidyverse', warn.conflicts = FALSE)
library('ggplot2', warn.conflicts = FALSE)
library('dplyr', warn.conflicts = FALSE)
library('corrplot', warn.conflicts = FALSE)
# Get this rmd file current absolute path 
# (from: https://stackoverflow.com/questions/47044068/get-the-path-of-current-script)
if (rstudioapi::isAvailable()) {
    wdir <- dirname(getActiveDocumentContext()$path)
}else{
    wdir <- getwd()
}
tryCatch(expr = {setwd(wdir)}, 
         error = function(e){print(e)}, 
         warning = function(w){print(w)})
```


## REFERENCES

List of references:

-   Source data [KDD Cup 1999 Data Data Set](https://archive.ics.uci.edu/ml/datasets/kdd+cup+1999+data)
-   A Detailed [Analysis of the KDD CUP 99 Data Set](https://www.ecb.torontomu.ca/~bagheri/papers/cisda.pdf)
-   KDD [download archive](https://www.kdd.org/kdd-cup/view/kdd-cup-1999/Data)
-   Kaggle comunity [notebooks](https://www.kaggle.com/datasets/galaxyh/kdd-cup-1999-data/code) with KDD CUP 99 data set.


## INITIAL ACTIONS

Based on the references we adopt the idea of using of two "reference" or "metadata" files anticpating analysis.
  1. A dataset with a categorization of the response variable values, the attack types (https://kdd.org/cupfiles/KDDCupData/1999/training_attack_types)
  2. A reference table with for each of the main data set features: a name, data type, stastictical type and description.


```{r read_data}
readingdataOK <- TRUE
tryCatch(
    expr = {
        # Loading the full dataset
        data_full <- read_csv("Book1.csv",
                          col_types = cols(SrcBytes = col_integer(),
                                           DstBytes = col_integer(), Land = col_integer(),
                                           WrongFragment = col_integer(), Urgent = col_number(),
                                           Hot = col_number(), NumFailedLogin = col_integer()))
        
        # Loading the training dataset
        data <- read.csv (file="Book2.csv",header=T)
        
        # Loading a reference data set with for each data feature: a name, data type, stastictical type and description
        # This info was collected from: 
        # - the header record for the name, 
        # - the R command str(data) for the data type, 
        # - the kddcup.names (http://kdd.org/cupfiles/KDDCupData/1999/kddcup.names) for the statistical type
        # - and asking chatgpt for a brief functional description for each field
        features <- read.csv (file="features.csv",header=T)
        nfeatures <- nrow(features)
        
        # Loading a data set with a categorization of the response variable: attack types 
        # (https://kdd.org/cupfiles/KDDCupData/1999/training_attack_types)        
        attackcats <- read.csv (file="attack_types.csv",header=F, col.names = c("Attack", "Category"))
        
        # Addition of response variable features based on the data$Attack values
        data$AttackCat <- apply(data, 1, function(row){
            category <- attackcats$Category[attackcats$Attack == row['Attack']]
            if(row['Attack']=='normal.'){
                category <- 'normal'
            }else if (length(category) == 0) {
                category <- "unknown"
            }else if (length(category) > 1) {
                category <- category[1]
            }
            return(category)
        })
        data$IsAttack <- apply(data, 1, function(row){
            isattack <- 0
            if(row['Attack']!='normal.'){
                isattack <- 1
            }
            return(isattack)
        })
            
        knitr::kable(features)

    }, error = function(e){
        readingdataOK <- FALSE
        print(e)
    }, warning = function(w){
        print(w)
    }
)
```

## Trying to understand correlations between variables

To find out correlations between variables, we used the cor() function (based on advice from chatGPT answering the question: <i>"How to find out about correlations between variables?"</i>).
  - The cor(data.frame()) expression calculates correlation coefficients which measure the linear association between columns of a data frame. 
  - The correlation coefficients range from -1 to +1, where -1 indicates a perfect negative linear relationship, +1 indicates a perfect positive linear relationship, and 0 indicates no linear relationship.
  - The resulting correlation matrix is a square matrix where each element represents the correlation between two variables.
  - For easiness of use we transformed the matrix into another data frame which we then order to get the highest correlations.

```{r raw_correlations}
corrmatrix <- cor(select_if(data, is.numeric))
corrmatrixcols <- colnames(corr_matrix)
corrmatrixrows <- rownames(corr_matrix)
corrs <- data.frame(feature1 = c(), feature2 = c(), coefficient = c())
for(c in 1:ncol(corrmatrix)){
    for(r in c:nrow(corrmatrix)){
        if(corrmatrixcols[c] != corrmatrixrows[r]){
            corrs <- rbind(corrs
                           , data.frame(feature1 = corrmatrixcols[c]
                                        , feature2 = corrmatrixrows[r]
                                        , coefficient = round(corrmatrix[r, c], digits = 5)))
        }
    }
}
corrs <- corrs[order(desc(corrs$coefficient)), ]

knitr::kable(corrs)
```

## Exploration of the training dataset

```{r data_exploration}
ncols <- ncol(data)
if(ncols != nfeatures){
    cat(paste("the number of cols on the trainig dataset,", ncols, 
              "does not match the expected number of features", nfeatures, "."))
    readingdataOK <- FALSE
}
if(readingdataOK)
{
    respvarAttack <- data[, 42]
    respvarAttackcat <- data[, 43]
    respvarIsAttack <- data[, 44]
    for(i in 1:ncols) {
        df <- data.frame(data[ , i], data$Attack, data$AttackCat, data$IsAttack)
        colnames(df) <- c(features[i,1], 'Attack', 'AttackCat', 'IsAttack')
        minval <- min(df[ , 1])
        maxval <- max(df[ , 1])
        cat(paste('----------------------------\nDetails of feature', i, '[', features[i,1], ']:\n'), fill = TRUE)
        cat(paste('   Data type:', features[i, 2]), fill = TRUE)
        cat(paste('   Statistical type:', features[i, 3]), fill = TRUE)
        cat(paste('   Field Description:', features[i, 4], '\n'), fill = TRUE)
        nmissingvals <- sum(is.na(df[ , 1])) # 'is.finite()' function is used to identify non-NA and non-NaN values
        cat(paste('   Number of Missing values:', nmissingvals, '\n\n'), fill = TRUE)
        if(minval == maxval){ 
            print('Both minimum and maximum values are the same!')            
        }else {
            cat(paste('   Value Minimum:[', minval, '], Value Maximum:[', maxval, ']'), fill = TRUE)
            if (features[i, 3] == 'symbolic'){
                values <- levels(factor(df[ , 1])) # Getting the values
                values.dis <- data.frame(table(df[ , 1])) # Getting values distribution
                colnames(values.dis)[1] <- "Vals"
                colnames(values.dis)[2] <- "Freq"
                nvals <- nrow(values.dis)
                values.dis <- values.dis[order(values.dis$Freq, decreasing=TRUE),]
                if(nvals >= 30){
                    print(values.dis[1:30,], row.names = FALSE)
                    cat(paste("\n... [", (nvals-30), "more]" ), fill = TRUE)
                    barplot(values.dis$Freq[1:15], names.arg = values.dis$Vals[1:15], 
                            xlab = "Values", ylab = "Occurrences", main = features[i,1])
                }else{
                    print(values.dis, row.names = FALSE)
                    barplot(values.dis$Freq, names.arg = values.dis$Vals, 
                            xlab = "Values", ylab = "Occurrences", main = features[i,1])
                }
            }else{
                colmean <- mean(df[ , 1])
                stdev <- sd(df[ , 1])
                if(stdev == 0){
                    print('   Standard Deviation is 0 for the feauture.')
                }else{
                    interquartile <- IQR(df[ , 1])
                    cat(paste('   Interquartile difference:[', interquartile, ']'), fill = TRUE)
                    if(interquartile == 0){print('      Requires looking for skewness.')}
                    
                    corr_matrix <- cor(select_if(df, is.numeric)) 
                    print(corr_matrix)
                    plot <- corrplot(corr_matrix, method = 'number') # colorful number
                    print(plot)

                    print(str(df[ , 1]))
                    colsummary <- summary(df[ , 1]) 
                    print(colsummary)
                    hist(df[ , 1])
                }
            }
            if( i < 42){
                cat(paste('Scatter plot for', features[i,1], 'vs. Attack:'), fill = TRUE)
                plot <- ggplot(data = df) + geom_point(mapping = aes(x = df[ , 1], y = Attack))
                print(plot)
                cat(paste('Scatter plot for', features[i,1], 'vs. Attack Categories:'), fill = TRUE)
                plot <- ggplot(data = df) + geom_point(mapping = aes(x = df[ , 1], y = AttackCat))
                print(plot)
                cat(paste('Scatter plot for', features[i,1], 'vs. Is Attack Indicator:'), fill = TRUE)
                plot <- ggplot(data = df) + geom_point(mapping = aes(x = df[ , 1], y = IsAttack))
                print(plot)
            }
        }
    }
}
```

